---
title: "cameraTrapAudioMothData"
author: "Nico Jaworski"
format: html
editor: visual
---

## TBS Species Distribution Based on Camera Trap Sightings

How species respond to risk is in part determined by the distribution of predators within their environment. Spatial heterogeneity in the distribution of predation causes variation in how animals perceive risk across the landscape. I deployed a grid of camera traps paired with autonomous acoustic monitoring devices at 20 monitoring stations to investigate the distribution of terrestrial predators across the landscape. In doing so, I aimed to asses if and how lowland woolly monkeys perceive and respond to different levels of risk across their environment. Monitor stations were placed at roughly 250 meter intervals from each other, with 5 stations running east to west and 4 running north to south, covering an area of an estimate 125 hectares.

Camera traps were able to document the presence of terrestrial predators and numerous prey species. Habitual use of particular locations by prey can be used as an indicator of potentially "safe" areas where risk is perceived as low.

This document displays the distribution of camera trap sightings at three levels: (1) the number of times a species was captures at a monitor station (2) the number of sightings of a species each hour per day and (3) the number of days a monitor station captured a species at least once. I constructed heat maps based on the rate of captures of each species to identify potential hot spots of activity across the landscape.

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(hms)
library(ggplot2)
library(sf)
library(leaflet)
library(leaflet.extras)
library(htmltools)

#Reading in number of sightings data, waypoint data, joinging####
f <- "https://raw.githubusercontent.com/NicoJaws23/TBS2026/refs/heads/main/RecordOfSightings.csv"
df <- read.csv(f)

g <- "https://raw.githubusercontent.com/NicoJaws23/TBS2026/refs/heads/main/RealMonWaypoints.csv"
wp <- read.csv(g)

df$Time <- as_hms(paste0(df$Time, ":00"))

df <- df|>
  mutate(NumberOfIndividuals = as.numeric(NumberOfIndividuals)) |>
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))

df_full <- left_join(df, wp, by = "MonitorID")

trails <- st_read("https://raw.githubusercontent.com/NicoJaws23/creative-data-visualization/refs/heads/main/TBS_Trails.geojson") |>
  st_transform(4326)
river <- st_read("https://raw.githubusercontent.com/NicoJaws23/creative-data-visualization/refs/heads/main/rio_tiputini.geojson") |>
  st_transform(4326)

north_arrow_html <- tags$div(
  style = "padding: 5px; background: rgba(255, 255, 255, 0.8); border-radius: 4px; text-align: center;",
  tags$div(style = "font-size: 20px; font-weight: bold; line-height: 1;", "▲"),
  tags$div(style = "font-size: 12px; font-weight: bold; font-family: sans-serif;", "N")
)

#Clean ScientificName

df$ScientificName <- trimws(df$ScientificName)  # removes leading/trailing spaces (fixes Leptotilla)

df$ScientificName <- recode(df$ScientificName,
                            # Capitalisation fixes
                            "Leptotilla Rufaxilla"     = "Leptotilla rufaxilla",
                            "Psophia Leucoptera"       = "Psophia leucoptera",
                            "Puma Concolor"            = "Puma concolor",
                            "Eira Barbara"             = "Eira barbara",
                            "Nasua Nasua"              = "Nasua nasua",
                            "Dasypus Novemcinctus"     = "Dasypus novemcinctus",
                            
                            # Typo fixes
                            "Lepardus paradalis"       = "Leopardus pardalis",
                            "Tianmus guttatus"         = "Tinamus guttatus",
                            "Mazama american"          = "Mazama americana",
                            "Myopracta pratti"         = "Myoprocta pratti",
                            "Cryptullerus soui"        = "Crypturellus soui"
)

df_species <- df[df$ScientificName != "N/A", ]

df_full <- left_join(df, wp, by = "MonitorID")
```

## Total Captures per Species

```{r}
#| echo: false
#| message: false
#| warning: false
species_summary <- df_species %>%
  group_by(MonitorID, ScientificName) %>%
  summarise(TotalDetections = n(), .groups = "drop")

species_summary <- left_join(species_summary, wp, by = "MonitorID")

ggplot(species_summary, mapping = aes(x = ScientificName, y = TotalDetections)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Heatmaps with leaflet#

expand_for_heatmap_1 <- function(data) {
  data[rep(seq_len(nrow(data)), times = data$TotalDetections), ]
}

species_list <- sort(unique(species_summary$ScientificName))

color_palette <- colorFactor(
  palette = "Set3",       # try also "Set1", "Set2", "Dark2", or "Paired"
  domain  = species_list
)

map_1Title<- tags$div(
  tag.map.title <- tags$style(HTML(".leaflet-control.map-title {
                                   background: rgba(255, 255, 255, 0.9);
                                   padding: 10px;
                                   font-weight: bold;
                                   font_size: 20px
                                   border-radius: 5px;
                                   }")),
  HTML("Total Number of Detections Per Species")
)
map_1 <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") |>
  addProviderTiles(providers$OpenStreetMap,     group = "Street Map") |>
  
  # ── Static layers ────────────────────────────────────────────────────────────
  addPolylines(
    data   = trails,
    color  = "black",
    weight = 2,
    group  = "Trails",
    label  = "Trails"
  ) |>
  addPolylines(
    data   = river,
    color  = "blue",
    weight = 2,
    group  = "River",
    label  = "River"
  ) |>
  addControl(
    html = map_1Title,
    position = "topright",
    className = "map-title"
  ) |>
  addControl(
    html = north_arrow_html,
    position = "topright"
  ) |>
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(metric = TRUE, imperial = FALSE)
  ) |>
  addLayersControl(
    baseGroups    = c("Satellite", "Street Map"),
    overlayGroups = c("Trails", "River", species_list),
    options       = layersControlOptions(collapsed = TRUE)
  )

for (sp in species_list) {
  
  sp_data     <- species_summary %>% filter(ScientificName == sp)
  sp_expanded <- expand_for_heatmap_1(sp_data)
  sp_color    <- color_palette(sp)   # unique color for this species
  
  map_1 <- map_1 %>%
    addHeatmap(
      data   = sp_expanded,
      lat    = ~lat,
      lng    = ~lon,
      blur   = 30,
      max    = 1,
      radius = 25,
      group  = sp
    ) %>%
    addCircleMarkers(
      data        = sp_data,
      lat         = ~lat,
      lng         = ~lon,
      radius      = ~scales::rescale(TotalDetections, to = c(3, 12)),
      color       = "white",
      fillColor   = sp_color,
      fillOpacity = 0.8,
      weight      = 1,
      popup       = ~paste0(
        "<b>", ScientificName, "</b><br>",
        "Monitor: ", MonitorID, "<br>",
        "Detections: ", TotalDetections
      ),
      group = sp
    )
}

map_1 <- map_1 %>% hideGroup(species_list)

map_1
```

## Total Captures Per Hour Per Day

```{r}
#| echo: false
#| message: false
#| warning: false
df_hourly <- df_species |>
  mutate(Hour = format(as.POSIXct(paste("2000-01-01", Time), format = "%Y-%m-%d %H:%M"), "%H")) |>
  distinct(MonitorID, ScientificName, Date, Hour, .keep_all = TRUE)

hourly_summary <- df_hourly |>
  group_by(MonitorID, ScientificName, Date) |>
  summarise(TotalDetections = n(), .groups = "drop")

hourly_map_summary <- hourly_summary |>
  group_by(MonitorID, ScientificName) |>
  summarise(TotalDetections = sum(TotalDetections), .groups = "drop")

hourly_map_summary <- left_join(hourly_map_summary, wp, by = "MonitorID")

ggplot(hourly_summary, mapping = aes(x = ScientificName, y = TotalDetections)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

species_list_3 <- sort(unique(hourly_summary$ScientificName))

color_palette <- colorFactor(
  palette = "Set3",       # try also "Set1", "Set2", "Dark2", or "Paired"
  domain  = species_list_3
)

map_3Title<- tags$div(
  tag.map.title <- tags$style(HTML(".leaflet-control.map-title {
                                   background: rgba(255, 255, 255, 0.9);
                                   padding: 10px;
                                   font-weight: bold;
                                   font_size: 20px
                                   border-radius: 5px;
                                   }")),
  HTML("Total Number of Detections Per Species Per Day Per Hour Per Monitor")
)

map_3 <- leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") |>
  addProviderTiles(providers$OpenStreetMap,     group = "Street Map") |>
  
  # ── Static layers ────────────────────────────────────────────────────────────
  addPolylines(
    data   = trails,
    color  = "black",
    weight = 2,
    group  = "Trails",
    label  = "Trails"
  ) |>
  addPolylines(
    data   = river,
    color  = "blue",
    weight = 2,
    group  = "River",
    label  = "River"
  ) |>
  addControl(
    html = map_3Title,
    position = "topright",
    className = "map-title"
  ) |>
  addControl(
    html = north_arrow_html,
    position = "topright"
  ) |>
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(metric = TRUE, imperial = FALSE)
  ) |>
  addLayersControl(
    baseGroups    = c("Satellite", "Street Map"),
    overlayGroups = c("Trails", "River", species_list),
    options       = layersControlOptions(collapsed = TRUE)
  )

for (sp in species_list_3) {   # also fix: use species_list_3, not species_list
  
  sp_data     <- hourly_map_summary %>% filter(ScientificName == sp)
  sp_expanded <- sp_data[rep(seq_len(nrow(sp_data)), times = sp_data$TotalDetections), ]
  sp_color    <- color_palette(sp)
  
  map_3 <- map_3 %>%
    addHeatmap(
      data   = sp_expanded,
      lat    = ~lat,
      lng    = ~lon,
      blur   = 30,
      max    = 1,
      radius = 25,
      group  = sp
    ) %>%
    addCircleMarkers(
      data        = sp_data,
      lat         = ~lat,
      lng         = ~lon,
      radius      = ~scales::rescale(TotalDetections, to = c(3, 12)),
      color       = "white",
      fillColor   = sp_color,
      fillOpacity = 0.8,
      weight      = 1,
      popup       = ~paste0(
        "<b>", ScientificName, "</b><br>",
        "Monitor: ", MonitorID, "<br>",
        "Detections: ", TotalDetections
      ),
      group = sp
    )
}

map_3 <- map_3 %>% hideGroup(species_list)

map_3
```

## Total Daily Captures Per Monitor

```{r}
#| echo: false
#| message: false
#| warning: false
df_uniqueSite <- df_full |>
  distinct(MonitorID, Date, ScientificName, .keep_all = TRUE) |>
  filter(ScientificName != "N/A")

uniqueSiteSummary <- df_uniqueSite |>
  group_by(MonitorID, ScientificName) |>
  summarise(TotalDetectionDays = n(), .groups = "drop")

uniqueSiteSummary <- left_join(uniqueSiteSummary, wp, by = "MonitorID")

ggplot(uniqueSiteSummary, mapping = aes(x = ScientificName, y = TotalDetectionDays)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

species_list_2 <- sort(unique(uniqueSiteSummary$ScientificName))

color_palette <- colorFactor(
  palette = "Set3",       # try also "Set1", "Set2", "Dark2", or "Paired"
  domain  = species_list_2
)

expand_for_heatmap_2 <- function(data) {
  data[rep(seq_len(nrow(data)), times = data$TotalDetectionDays), ]
}

map_2Title<- tags$div(
  tag.map.title <- tags$style(HTML(".leaflet-control.map-title {
                                   background: rgba(255, 255, 255, 0.9);
                                   padding: 10px;
                                   font-weight: bold;
                                   font_size: 20px
                                   border-radius: 5px;
                                   }")),
  HTML("Total Number of Detections Per Species Per Day Per Monitor")
)

map_2 <- leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") |>
  addProviderTiles(providers$OpenStreetMap,     group = "Street Map") |>
  
  # ── Static layers ────────────────────────────────────────────────────────────
  addPolylines(
    data   = trails,
    color  = "black",
    weight = 2,
    group  = "Trails",
    label  = "Trails"
  ) |>
  addPolylines(
    data   = river,
    color  = "blue",
    weight = 2,
    group  = "River",
    label  = "River"
  ) |>
  addControl(
    html = map_2Title,
    position = "topright",
    className = "map-title"
  ) |>
  addControl(
    html = north_arrow_html,
    position = "topright"
  ) |>
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(metric = TRUE, imperial = FALSE)
  ) |>
  addLayersControl(
    baseGroups    = c("Satellite", "Street Map"),
    overlayGroups = c("Trails", "River", species_list),
    options       = layersControlOptions(collapsed = TRUE)
  )

for (sp in species_list) {
  
  sp_data     <- uniqueSiteSummary %>% filter(ScientificName == sp)
  sp_expanded <- expand_for_heatmap_2(sp_data)
  sp_color    <- color_palette(sp)   # unique color for this species
  
  map_2 <- map_2 %>%
    addHeatmap(
      data   = sp_expanded,
      lat    = ~lat,
      lng    = ~lon,
      blur   = 30,
      max    = 1,
      radius = 25,
      group  = sp
    ) %>%
    addCircleMarkers(
      data        = sp_data,
      lat         = ~lat,
      lng         = ~lon,
      radius      = ~scales::rescale(TotalDetectionDays, to = c(3, 12)),
      color       = "white",
      fillColor   = sp_color,
      fillOpacity = 0.8,
      weight      = 1,
      popup       = ~paste0(
        "<b>", ScientificName, "</b><br>",
        "Monitor: ", MonitorID, "<br>",
        "Detection Days: ", TotalDetectionDays
      ),
      group = sp
    )
}

map_2 <- map_2 %>% hideGroup(species_list)

map_2
```
